Classless Java

### 2016-03-23 OOPSLA16/sections/RelatedWork.tex
Text in Section 4. Code style.

\section{Related Work}\label{sec:related}
In this section we discuss related work and how it compares to Classless Java.

@@

\paragraph{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favor or against multiple inheritance.  Multiple
inheritance is very expressive, but difficult to model and implement, and can
cause difficulty (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid models
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; the mixin model adds some
restrictions; and the trait model is the most restricted one (excluding
state, instantiation, etc).

@@

%\paragraph{Mixins.}
Mixins are a more restricted model than the C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearization: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearization (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearized mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearized mixin composition. % Other languages, such as
% Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearized mixins. Indeed, in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@
\end{comment}
\noindent Java interfaces and default methods do not use
linearization: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

%\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which
aims to do multiple inheritance without the diamond problem. They
divide inheritance into two separate concepts: inheritance dependency
and implementation inheritance. Using a combination of
\texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover,
fields and multiple inheritance can coexist. However untangling
inheritance also untangles the class structure. In CZ, not only the
number of classes, but also the class hierarchy complexity
increases. \mixin does not complicate the class structure, and state
can also coexist with multiple inheritance.

%\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality, ignoring state and state
initialization. Classes, as object factories, require functionality from
(multiple) traits. Java 8 interfaces are closely related to
traits: concrete method implementations are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavors of multiple inheritance in Java. Traits offer an algebra
of composition operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.  We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The first author of~\cite{bono14} agreed (via personal communication) that
our revised version for exclusion is cleaner, typesafe and more direct.
\newcommand\shortItem{\vspace{-1ex}}
\begin{itemize}
\item \textbf{Symmetric sum} can be obtained by simple multiple inheritance between interfaces.
    \begin{lstlisting}
    interface A { int x(); }    interface B { int y(); }    interface C extends A, B {}
    \end{lstlisting}
\shortItem
\item \textbf{Overriding} a conflict is obtained by specifying which super interface take precedence.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B { default int m() {return 2;} }
    interface C extends A, B { default int m() {return B.super.m();} }
    \end{lstlisting}
\shortItem
\item \textbf{Alias} is creating  a new method delegating to the existing super interface.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { default int k() {return A.super.m();} }
    \end{lstlisting}
\shortItem

\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\shortItem
\end{itemize}

@@

% \footnote{The original
%  notion of traits by Scharli et al. prescribes, among other things,
%  that: 1) a trait provides a set of methods that implement behavior;
%  and 2) a trait does not specify any state variables, so the methods
%  provided by traits do not access state variables directly. Java 8
%  interfaces follow similar principles too. Indeed, a detailed
%  description of how to emulate trait-oriented programming in Java 8
%  can be found in the work by Bono et al.~\cite{bono14}. The Java 8
%  team designing default methods, was also fully aware of that
%  secondary use of interfaces, but it was not their objective to model
%  traits: ``The key goal of adding default methods to Java was
%  "interface evolution", not "poor man's
%  traits"''~\cite{goetz13default}. As a result, they were happy to
%  support the secondary use of interfaces with default methods as long
%  as it did not make the implementation and language more complex.}
\marcoT{We could use Scala Traits to encode IB, but the simpler Java8
  model is a better starting point: scala traits can declare \Q@def, val, var@ and
  \Q@lazy val@ However, while \Q@def@s can be overridden by
  \Q@val@/\Q@var@s, \Q@val@/\Q@var@s can not be overridden.
   A\Q@val@/\Q@var@ member is predestined to be
  instance state, and this is in sharp contrast with our model where
  the state is only a private detail of instances, and do not
  contaminate interface/trait/classes.
  }

@@

\paragraph{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}
In certain situations, object interfaces allow automatic refinement for \emph{return
types}. This is part of a bigger topic in class-based languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside of the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But our approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Nevertheless, as illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications.
% The addition of MyType to a language will allow easy definition of
%binary methods, methods with recursive types (i.e., the same type of the
%receiver appears in the argument or return positions of methods), etc.

>>>>

\section{Related Work}\label{sec:related}
%In this section we discuss related work and how it compares to Classless Java.

@@

\paragraph{Models of Object-Oriented Programming}
There are traditionally two big styles of OOP: class-based and
prototype-based. The class-based tradition started with Simula~\cite{Dahl66simula}, 
and is the dominant style of OOP today. Major languages such 
as Java, C\#, C++ or Scala favour a class based OOP style. Very often
class-based languages are statically typed and they have interfaces or
some other similar mechanisms such as traits. However, in contrast to IB
programming, interfaces or traits in CB programming are not meant to
replace classes, but rather, they complement classes. Normally
interfaces or traits define the signatures of the methods and
\emph{may} have some behaviour (such as in Java8 interfaces or Scala
traits). However classes are still responsible for object
instantiation and handling state. Prototype-based languages started
with Self~\cite{Ungar87self}, and are still widely used today via languages like
JavaScript. Frequentely prototype-based languages are dynamically 
typed. IB programming is clearly closer to CB programming. However the
key difference is that classes are not needed in IB
programming. Object interfaces recover the responsability of object
instantiation, and abstract state operations provide a way to deal
with state. A primary motivation of IB programming is to improve 
modularity and flexibility. Central to these goals are the 
more expressive and automated forms of type-refinement, and support 
for multiple interface inheritance.

\paragraph{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favor or against multiple inheritance.
Multiple \emph{implementation}
inheritance is very expressive, but difficult to model and implement, and can
cause difficulties (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid models
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; the mixin model adds some
restrictions; and the trait model is the most restricted one (excluding
state, instantiation, etc).

@@

%\paragraph{Mixins.}
Mixins are more restricted than the C++ approach. Mixins allow naming
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearization: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearization (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearized mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as the diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearized mixin composition. % Other languages, such as
% Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearized mixins. Indeed, in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@
\end{comment}
\noindent Java interfaces and default methods do not use
linearization: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

%\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which
aims to do multiple inheritance without the diamond problem. They
divide inheritance into two separate concepts: inheritance dependency
and implementation inheritance. Using a combination of
\texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover,
fields and multiple inheritance can coexist. However untangling
inheritance also untangles the class structure. In CZ, not only the
number of classes, but also the class hierarchy complexity
increases. IB programming does not complicate the hierarchical structure, and state
can also coexist with multiple inheritance.

%\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality, ignoring state and state
initialization. Classes, as object factories, require functionality from
(multiple) traits. Java 8 interfaces are closely related to
traits: concrete method implementations are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavors of multiple inheritance in Java. Traits offer an algebra
of composition operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.

\begin{comment}
We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The first author of~\cite{bono14} agreed (via personal communication) that
our revised version for exclusion is cleaner, typesafe and more direct.
\newcommand\shortItem{\vspace{-1ex}}
\begin{itemize}
\item \textbf{Symmetric sum} can be obtained by simple multiple inheritance between interfaces.
    \begin{lstlisting}
    interface A { int x(); }
    interface B { int y(); }
    interface C extends A, B {}
    \end{lstlisting}
\shortItem
\item \textbf{Overriding} a conflict is obtained by specifying which super interface take precedence.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B { default int m() {return 2;} }
    interface C extends A, B {
        default int m() {return B.super.m();} }
    \end{lstlisting}
\shortItem
\item \textbf{Alias} is creating  a new method delegating to the existing super interface.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A {
        default int k() {return A.super.m();} }
    \end{lstlisting}
\shortItem

\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\shortItem
\end{itemize}
\end{comment}

@@

% \footnote{The original
%  notion of traits by Scharli et al. prescribes, among other things,
%  that: 1) a trait provides a set of methods that implement behavior;
%  and 2) a trait does not specify any state variables, so the methods
%  provided by traits do not access state variables directly. Java 8
%  interfaces follow similar principles too. Indeed, a detailed
%  description of how to emulate trait-oriented programming in Java 8
%  can be found in the work by Bono et al.~\cite{bono14}. The Java 8
%  team designing default methods, was also fully aware of that
%  secondary use of interfaces, but it was not their objective to model
%  traits: ``The key goal of adding default methods to Java was
%  "interface evolution", not "poor man's
%  traits"''~\cite{goetz13default}. As a result, they were happy to
%  support the secondary use of interfaces with default methods as long
%  as it did not make the implementation and language more complex.}

\begin{comment}
\marcoT{We could use Scala Traits to encode IB, but the simpler Java8
  model is a better starting point: scala traits can declare \Q@def, val, var@ and
  \Q@lazy val@ However, while \Q@def@s can be overridden by
  \Q@val@/\Q@var@s, \Q@val@/\Q@var@s can not be overridden.
   A\Q@val@/\Q@var@ member is predestined to be
  instance state, and this is in sharp contrast with our model where
  the state is only a private detail of instances, and do not
  contaminate interface/trait/classes.
  }
\end{comment}

@@

\paragraph{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}
Object interfaces support automatic type-refinement.
Type refinement is part of a bigger topic in class-based languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside of the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional 
complexity from the type system.

% The addition of MyType to a language will allow easy definition of
%binary methods, methods with recursive types (i.e., the same type of the
%receiver appears in the argument or return positions of methods), etc.
